---
title: "Snow Detection and Fractional Snow Cover with snowsense"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Snow Detection and Fractional Snow Cover with snowsense}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
This vignette demonstrates a full snow analysis workflow using a Sentinel-2 L1C scene (Stacked to a single tif, resampled to 20m) from the Austrian Alps (June 2023). 
The corresponding Github Repository can be found [here](https://github.com/Lutzkk/snowsense), which also contains the dataset shown here. 
The code for the plots is left out in this vignette.

The workflow includes:

1. Data Loading
2. Snow Detection 
3. Despeckling 
4. Fractional Snow Cover Estimation using two different regression approaches

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(terra)
library(tidyterra)
library(ggplot2)
library(snowsense)
library(patchwork)
```

## Data Loading
The example dataset is a clipped Sentinel-2 L1C scene (dtype: **uint16**, bands: **B01 - B12**) 
which has been stacked to a single multi-band GeoTIFF and resampled to 20m resolution.



```{r}
# Load Packages 
library(snowsense)

ds <- rast("../inst/extdata/example_s2_20m.tif") #load sentinel2 data as SpatRaster with terra
ds <- ds / 10000 #scale to reflectance values (0-1)

print(paste("Bands available:", paste(names(ds), collapse = ", ")))

plotRGB(ds, r="B04", g="B03", b="B02", stretch="lin", main="Sentinel-2 L1C RGB Scene (20m, Stacked)") 

```

## Snow Detection
`snowsense` supports three spectral indices for snow detection (**NDSI**, **RGB_brightness**, and **[WSI](https://doi.org/10.1007/s00703-020-00749-y)**). 
NDSI is demonstrated here as it suits the Sentinel-2 data best. An approach with a fixed threshold (0.4) and an unsupervised approach using
[Otsu's method](https://doi.org/10.1109/TSMC.1979.4310076) are shown for comparison.
```{r ndsi, fig.width=10, fig.height=5}
result_ndsi <- detect_snow(ds, index = "ndsi",
                           bands = list(green = "B03", swir = "B11"),
                           threshold = 0.4)

# Note that if no threshold is set, the function falls back to Otsu's method. An unsupervised
# approach that selects an optimal threshold for separating bimodal grayscale data into two classes
result_ndsi_otsu <- detect_snow(ds, index = "ndsi",
                                bands = list(green = "B03", swir = "B11"),
                                threshold = NULL)
```


```{r, ndsi_plot, echo=FALSE, fig.width=10, fig.height=8}
# Prepare snow masks (0 -> NA for transparency)
mask_thresh <- result_ndsi$binary_mask
mask_thresh[mask_thresh == 0] <- NA
mask_thresh <- as.factor(mask_thresh)

mask_otsu <- result_ndsi_otsu$binary_mask
mask_otsu[mask_otsu == 0] <- NA
mask_otsu <- as.factor(mask_otsu)

no_axes <- theme(axis.text = element_blank(), axis.ticks = element_blank(),
                 plot.margin = margin(2, 2, 2, 2))

rgb_layer <- geom_spatraster_rgb(data = ds, r = 4, g = 3, b = 2, stretch = "lin")

p1 <- ggplot() + rgb_layer + theme_minimal() + no_axes +
  labs(title = "RGB")

p2 <- ggplot() + geom_spatraster(data = result_ndsi$index) +
  scale_fill_viridis_c(name = "NDSI", na.value = NA) +
  theme_minimal() + no_axes +
  labs(title = "NDSI Values")

p3 <- ggplot() + rgb_layer +
  geom_spatraster(data = mask_thresh) +
  scale_fill_manual(values = c("1" = "cyan"), na.value = NA, na.translate = FALSE, guide = "none") +
  theme_minimal() + no_axes +
  labs(title = "Snow Mask (threshold = 0.4)")

p4 <- ggplot() + rgb_layer +
  geom_spatraster(data = mask_otsu) +
  scale_fill_manual(values = c("1" = "cyan"), na.value = NA, na.translate = FALSE, guide = "none") +
  theme_minimal() + no_axes +
  labs(title = paste0("Snow Mask (Otsu = ", round(result_ndsi_otsu$threshold, 2), ")"))

(p1 | p2) / (p3 | p4) 
```
The bottom right plot depicts the snow mask derived from using Otsu's method for thresholding. The automatically selected lower threshold here works better, capturing more snow pixels and reducing false negatives from the fixed threshold (from a visual inspection). However, with a fixed threshold you have full control and reproducibility across scenes. RGB_brightness is especially useful for high resolution drone data, where only RGB is available.
WSI is based on a paper from Donmez et al. (2021) and is especially good at discriminating between water and snow.

## Despeckling

Snow masks often contain speckle noise (isolated snow classified pixels that may be noise or small patches that can be caused by misclassification). `despeckle_snow()` removes these using majority filtering and/or a minimum patch size filter.

```{r despeckle}
clean <- despeckle_snow(result_ndsi_otsu$binary_mask, window=9, min_pixels=5)
```

```{r, despeckle_plot, echo=FALSE, fig.width=10, fig.height=4}
mask_clean <- clean
mask_clean[mask_clean == 0] <- NA
mask_clean <- as.factor(mask_clean)

p_before <- ggplot() + rgb_layer +
  geom_spatraster(data = mask_otsu) +
  scale_fill_manual(values = c("1" = "cyan"), na.value = NA, na.translate = FALSE, guide = "none") +
  theme_minimal() + no_axes +
  labs(title = "Before Despeckling")

p_after <- ggplot() + rgb_layer +
  geom_spatraster(data = mask_clean) +
  scale_fill_manual(values = c("1" = "cyan"), na.value = NA, na.translate = FALSE, guide = "none") +
  theme_minimal() + no_axes +
  labs(title = "After Despeckling (window=9, min_pixels=5)")

p_before | p_after
```
The despeckling function removes small isolated snow pixels that are likely false positives. The parameters `window` and `min_pixels` control the size of the neighborhood (majority based filtering) and the minimum number of connected pixels required to retain a snow patch (size based filtering).
The plot shows that after despeckling a lot of tiny blue dots have been removed, indicating it is an effective way to clean up the snow mask from tiny positives. However, it also removes some small snow patches that may be real, so there is a tradeoff between removing noise and losing small features.

## Fractional Snow Cover Estimation

Rather than a binary snow mask, `snow_cover_fraction()` estimates the fraction of snow cover per pixel directly using NDSI-based regression. This is useful when sub-pixel snow cover information is needed, for example at coarser resolutions.

Two regression approaches are currently supported in the package:
- **[Salomonson & Appel (2004)](https://doi.org/10.1016/j.rse.2003.10.016)**: FSC = 1.45 * NDSI - 0.45
- **[Gascoin et al. (2020)](https://doi.org/10.3390/rs12182904)**: FSC = 0.5 *  tanh(2.65 * NDSI - 1.42) + 0.5

```{r, fsc}
fsc_sal <- snow_cover_fraction(ds, bands = list(green = "B03", swir = "B11"), method = "salomonson")
fsc_sig <- snow_cover_fraction(ds, bands = list(green = "B03", swir = "B11"), method = "sigmoid")
```

```{r, fsc_plot, echo=FALSE, fig.width=10, fig.height=4}
fsc_sal_overlay <- fsc_sal
fsc_sal_overlay[fsc_sal_overlay < 0.1] <- NA

fsc_sig_overlay <- fsc_sig
fsc_sig_overlay[fsc_sig_overlay < 0.1] <- NA

p_sal <- ggplot() + rgb_layer +
  geom_spatraster(data = fsc_sal_overlay) +
  scale_fill_viridis_c(name = "FSC", limits = c(0.1, 1), na.value = NA) +
  theme_minimal() + no_axes +
  labs(title = "FSC (Salomonson & Appel, 2004)")

p_sig <- ggplot() + rgb_layer +
  geom_spatraster(data = fsc_sig_overlay) +
  scale_fill_viridis_c(name = "FSC", limits = c(0.1, 1), na.value = NA) +
  theme_minimal() + no_axes +
  labs(title = "FSC (Gascoin et al., 2020)")

p_sal | p_sig
```


